A large and variable leading tail of Helium in a sub-Saturn undergoing runaway inflation

### 2023-06-21 paper/paper1/ms.tex
Version 0.6 of paper destined for OVERLEAF.
Direct edits from Overleaf, including GS, and MM

\title{A large and variable leading tail of Helium in a sub-Saturn undergoing runaway inflation}

@@

A runaway inflation scenario predicts an inevitable and profound mass loss rate for inflated hot Saturns.  As the planet's atmosphere overflows its Roche lobe at an ever-increasing pace instantaneous mass loss rates may exceed $\dot{M}>10^{13}$ g/s, over 10$\times$ larger than those previously seen in atmospheric escape measurements to date \citep{2022arXiv221116243D}.  Inflated hot Saturns, therefore, make excellent targets for direct measurement of atmospheric escape.

Large mass loss alone does not guarantee detectability.  The ability to detect even immense mass loss hinges on its observability in spectral tracers.  Ly$\alpha$, \ion{He}{1} 10833 \AA, and H$\alpha$ have emerged as the most amenable to detection \citep{2018ApJ...855L..11O,2022arXiv221116243D,2023MNRAS.518.4357O}, but each of these has its own observational limitations.  Ly$\alpha$ can suffer from interstellar medium \ion{H}{1} absorption censoring its low-velocity line core, for example.  Here we focus on \ion{He}{1} 10833 \AA, which offers some advantages.  In particular, metastable Helium's ability to be observed at high spectral resolution from the ground has been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.  \ion{He}{1} has resulted in at least 14 systems with detections, including \object[GJ 3470]{GJ3470} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.  An observational picture of mass loss in hot Saturns appears to be lacking, at least in part due to their intrinsic rarity.

@@

Here we present a multi-year observational campaign searching for atmospheric escape in \object{HAT-P-67 b}, an inflated hot Saturn orbiting an F type sub-giant at an orbital separation of 0.06 AU and a 4.8-day orbital period \citep{2017AJ....153..211Z}.  Its strong insolation ($T_\mathrm{eq}\sim2000\;$K) combined with \object{HAT-P-67 b}'s extremely low surface gravity makes it an exceptional candidate for strong atmospheric mass loss via Roche lobe overflow.  Its $\sim$0.05 g~cm$^{-3}$ density places it firmly below the 0.1 g~cm$^{-3}$ threshold predicted to exhibit runaway inflation by \citet{2023ApJ...945L..36T}.  Figures \ref{fig:instellation} and \ref{fig:ThornFortUpdated} show how much of an outlier \hatpb~ is: large, low mass, and heavily irradiated.

The evolutionary state of \hatp offers even more intriguing possibilities.  The host star may be an evolved subgiant, opening the prospect for planetary ``re-inflation'' \citep{2022AJ....163...53S,2022AJ....163..120G,2023arXiv230306728G} as the planet's insolation increased with the star's rapid luminosity increase in this part of the short-lived HR diagram.  The tidal gravity of the nearby massive host star may amplify mass loss rates even further \citep{2007A&A...472..329E,2023ApJ...945L..36T}. Its status as a rare inflated Hot Saturn makes \object{HAT-P-67 b} a promising laboratory, uniquely suited for testing the runaway inflationary predictions of Ohmic dissipation and XUV photoevaporation.

We assemble both archival, previously published, and new observations of the HAT-P-67 system, chronicled in Section \ref{secObs}.  In Section \ref{secAnalysis} we refine the stellar and planet properties based on those observations, including distance, radius, and orbit re-analyses.  The Helium excess detection is presented in \ref{secHeAnalysis}, with deep dives on the signal's trend with orbital phase.  We use this orbital structure to reconstruct the geometry and mass loss of the escaping Helium exosphere, with both 1D and 3D models (Section \ref{secResults}).  We assess the physical mechanisms (Section \ref{secPhysMech}) giving rise to the escaping atmosphere, weighing the distinctive predictions of XUV irradation and Ohmic dissipation.  Finally in Section \ref{secDiscuss}, we question the assumptions in our approach, discuss the overall congruence of predictions and observations, and highlight some implications for future exosphere studies.

@@

The Habitable Zone Planet Finder Spectrograph \citep[HPF;][]{2012SPIE.8446E..1SM,2014SPIE.9147E..1GM, 2019Optic...6..233M} on the queue-scheduled 10-meter \emph{Hobby-Eberly Telescope} \citep[HET;][]{1998SPIE.3352...34R} operates in the near-IR from $8100-12800~$\AA~ spanning the \textit{z}, \textit{Y}, and \textit{J} bands at spectral resolving power $R=55,000$. The HET fixed-elevation design \citep{2007PASP..119..556S} limits the observability of \object{HAT-P-67} to less than 1 hour ``tracks'' for a fixed range of hour angles before (east track) and after (west track) the star transits the meridian.  Whereas conventional steerable telescopes could conduct continuous point-and-stare observations of \object{HAT-P-67} for hours, HET cannot.  In practice, this limitation means that in-transit and out-of-transit observational phases were rarely possible on the same night.  Instead, we organized the observations into four campaigns to coincide with \object{HAT-P-67 b} transits on the nights beginning 2020 April 27, 2020; May 21, 2020; June 14, 2020; and 2022 April 28.  These campaigns have out-of-transit observations at least one night before and one night after, and often two nights on either side of the transit.  Two more transit snapshots were obtained in 2022 June-July without the accompanying visits immediately before and after.  The in-transit campaigns had up to 14 exposures per HET track, with integration times between 5-8.5 minutes.  We also obtained random-in-phase low-priority ``P4'' queue-filler observations \citep{2007PASP..119..556S}.  These out-of-transit snapshot observations typically received 4 or fewer individual exposures.  We observed HAT-P-67 with HPF for a total of 41 visits on 39 unique nights, with two of those nights observing both the east and west HET tracks.  The total on-source integration time exceeds 13.8 hours, over $N=152$ individual spectra.  Table \ref{tabHPFLog} summarizes the log of observations.

@@

\subsection{TESS Light Curve}
Previously, \hatpb~transits had only been detected with \emph{HATNet} \citep{2004PASP..116..266B} and followed up with KeplerCam on the FLWO 1.2 m telescope \citep{2017AJ....153..211Z}. These ground-based photometers were not intended to measure long-term stellar variability signals.  We, therefore, examined the \emph{TESS} lightcurves for out-of-transit photometric variability.  The revised precision and continuous coverage of \emph{TESS} can refine the orbital solution.

@@

We reduced and analyzed the HPF data first for orbit fitting via precision radial velocity (PRV) measurements following the procedures described in \citet{2021AJ....161..173T} and updated for joint lightcurve and RV fits \citep{2022AJ....163..225T}.  Here we included the new TESS lightcurves and the existing Keck HIRES points reported in Table 3 of \citet{2017AJ....153..211Z}.  The HPF points exhibited an RV jitter of $\sim164\;\kms$, a few times larger than the typical Keck HIRES measurements.  So even though the HPF points were more numerous, their marginal value for RV orbit determination was subdued owing to the lower RV information content in the near-IR for this relatively rapidly rotating F star.  Figure \ref{fig:RVfit} shows the fitted RV semiamplitude $K=33_{-15}^{+21}$ \kms---assuming a circular orbit---slightly decreasing the uncertainty compared to the previous $1\sigma$ limit.  Table \ref{tabOrbit} lists the revised properties, with minor differences from the existing values.
\authorcomment1{Quang Tran: please verify and revise.}

@@

\subsection{Velocity substructure}
Figure \ref{fig:centroids} shows the centroid positions of the Helium excess feature.  The bulk velocity drifts of the planet's outflow can be seen with a few conspicuous trends.  The pre-transit phases ($-0.3$ to $+0.03$) show a slight tendency towards blueshift relative to stellar restframe, with an accelerating blueshift from phases $-0.3$ to $-0.15$, albeit with some visit-to-visit scatter.  The post-transit observations ($+0.03$ to $+0.015$) exhibit weak-albeit-significant equivalent widths, as denoted by their smaller marker size.  The centroids dramatically accelerate to larger bulk blueshifts, as low as 10831 \AA.

The densely sampled in-transit phases are shown as a zoomed-in, oversized inset.  Here we can see the planetary rest frame as the dotted line cutting across the panel.  At the moment of planet ingress, the centroids exhibit a redshifted absorption relative to the planet restframe. Measurements from April 2022 and May 2020 transits partially overlap in phase coverage while yielding slightly different line centroid locations.  Both visits appear consistent with near-zero restframe velocity.  The May 2020 visit shows a slight trend towards alignment with the planet restframe.  The April 2020 and June 2020 partial transits sampled phases close to planet egress, with a clear blueshift relative to both the star and planet.

@@

The predicted Helium ionization depends sensitively on the XUV flux \citep{2019ApJ...881..133O}.  Ideally, we would have a measurement of the XUV spectrum of HAT-P-67, but the limited facilities and distance of the source preclude these challenging observations.  Instead, we constructed panchromatic X-ray to visible SEDs by scaling and stitching together synthetic spectra to provide a range of high and low $L_X/L_\mathrm{bol}$.  We chose a 6400 K PHOENIX \citep{husser13} solar metallicity photospheric spectral template with $\log{g}=3.5$, scaled to the solid angle seen by \hatpb.  The F7IV-V \object[tau Boo]{$\tau$ Boo} serves as the closest analog with available synthetic X-ray coronal spectra \citep{2011A&A...532A...6S}.  But \hatp resides closer to the Kraft break than does \object[tau Boo]{$\tau$ Boo}, and the move to hotter F stars may be associated with weakening of the stellar wind and other atmospheric changes leading to the prospect of much lower XUV luminosity for HAT-P-67 \citep{2022ApJ...930....7A}.  So there remains significant uncertainty about the applicability of the \object[tau Boo]{$\tau$ Boo} XUV spectrum to HAT-P-67.  We scaled the synthetic SED of \object[tau Boo]{$\tau$ Boo} retrieved from X-exoplanets \citep{2011A&A...532A...6S} so that the integral of ground-state ionizing photons ($\lambda<504\AA$) exhibited $L_X/L_\mathrm{bol} \in [10^{-6}, 10^{-4}]$.  Figure \ref{fig:XUV} shows two conceivable SEDs with these high and low radiation hardnesses.

@@

Equipped with these two SEDs of differing radiation hardness, we explored different mass loss rates and exosphere temperatures.  Figure \ref{fig:pwinds} shows one example model spectrum with $\dot{M} = 10^{13}$ g/s, and $T_0=9100$~K, with an XUV spectrum possessing $L_x/L_\mathrm{bol}=10^{-5}$.  This mass loss rate would imply a characteristic lifetime of $\sim$2000 Myr.  The model spectrum exhibits approximately the same equivalent width as our HPF observations, making it a hypothetical scenario among a family of partially degnerate solutions.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/pwinds_data_overlay.png}
    \caption{Simulated 1D Parker wind model of Helium absorption in \hatpb.  The spectrum was generated with a mass loss rate of $10^{13}$ g/s, exosphere gas temperature of 9100 K, and with an XUV luminosity $L_x/L_\mathrm{bol}=10^{-5}$ intermediate between those shown in Figure \ref{fig:XUV}.  The \texttt{p-winds} model yields the \ion{He}{1} 10830 triplet lines (colored dashed lines), with a cumulative feature in black possessing about the same equivalent width as the data, shown as the in-transit mean.  Additional unmodeled velocity dispersion of the gas can explain differences between the line shapes of the model and data.  Overall the 1D models may be too simplistic to represent the inherently 3D structure of the outflow.}
    \label{fig:pwinds}
\end{figure}

@@

Orbtial shear, stellar wind confinement, and preferential mass loss may all consipire to create the conditions of high enough column density to populate the \ion{He}{1} metastable triplet to detectable levels while imbuing the velocity substructure that we see.  Here we interpret the bulk velocity substructure in Figure \ref{fig:centroids} under these gas dynamics mechanisms.  We attribute the pre-transit arc to stellar wind acceleration.  The stellar wind is initially too weak to plow the dense escaping gas until a separation of 60-130 planetary radii ($+0.15$ orbital phase), the inflection point in an enormous bow shock.  Gas at these separations has had enough time to diffuse, both out of the orbital plane and to slightly larger radii shells, in turn lowering the column density along the line-of-sight.  The lower column density provides both fewer Helium atoms to participate in absorption, and less overall NUV shielding, leading to greater fractional ionization and further depopulation of the \ion{He}{1} metastable state.

@@

\begin{equation}
    \dot{M} = \frac{3}{4}\frac{\eta F_\mathrm{XUV}}{G K_t \rho_\mathrm{XUV}}\label{thorn23Mdot}.
\end{equation}

\noindent where $K_t$ is the tidal gravity term, and $\eta \sim 0.4$ comes from \citet{2022A&A...663A.122C}.  Assuming $\log{L_\mathrm{XUV}/L_\mathrm{bol}}=-4.2$, we would expect \hatpb~ to exhibit $\dot{M}\sim10^2$ M$_\oplus$/Gyr. At a current $M\sim95\;$M$_\oplus$ and an accelerating mass loss rate, its lifetime would measure in the $\sim$100 Myr range.  The \citet{2023ApJ...945L..36T} simulations focused on 0.75-1.25 $M_\odot$ host stars; the higher mass $\sim$1.6 $M_\odot$ \hatp would deliver greater tidal gravity for a given semi-major axis, and possibly lower $\log{L_\mathrm{XUV}/L_\mathrm{bol}}$ compared to more active G stars.  Nevertheless, we can project the time series trends in their Figures 3 and 4 to recreate a qualitative history and fate of \hatpb~ under the assumptions of XUV photoevaporation.  Broadly, the planet would have started with an initial mass up to 50$\%$ greater than at present, with $R\sim\;$1.4 $R_\mathrm{Jup}$, for an initial density of $\sim$0.2 g/cm$^-3$.  It would lose mass at a rate of tens of Earth masses per Gyr for its 1 Gyr lifetime, expanding modestly until it reaches the critical $\sim$0.1 g/cm$^-3$ threshold, at which point the mass loss rate increases to its current value.  Having lost a significant portion of its mass, its orbit expands to conserve angular momentum.  It will last only tens of Myr in its current state before losing almost all of its envelope and settling as a $5-15$ M$_\oplus$ core with a final radius of 0.2$-$0.3 R$_\mathrm{Jup}$.

\subsection{Ohmic Dissipation-driven mass loss}
The Ohmic dissipation mechanism made two key observable predictions.  Heating efficiency should initially increase as a function of equilibrium temperature, then degrade as irradiation exceeds equilibrium temperatures of about 2000 K.  Second, hot Saturns ($\lesssim0.5 M_\mathrm{Jup}$) should undergo runaway evaporation, whereas hot Jupiters ($\gtrsim 1 M_\mathrm{Jup}$) should reach stable equilibrium---albeit inflated---radii after Gyr timescales.  Both of these outcomes predated data that could validate them, and these outcomes differ from the behavior of other heating mechanisms, such as photoevaporation or tides.

@@

A few caveats complicate the unambiguous causal interpretation of Ohmic dissipation.  First, the scaling law arguments that produced Equation \ref{eqInflate} were only proposed as coarse scaling law estimates, with numerical simulations needed to quantify inflation timescales for individual systems.  Accordingly, a $10\times$ higher inflation timescale of 50 Myr---allowed by the coarse scaling law---would yield $\dot{M}_\mathrm{infl}\sim10^{14}$ g/s, still a very large mass loss rate, and only a factor of 5 away from the baseline 1D model.  Factors of a few uncertainties in the Ohmic dissipation efficiency $\epsilon$ may also contribute.

@@

\citet{2017AJ....153..211Z} estimated that \hatpb~received about twice the incident flux as a Zero Age Main Sequence (ZAMS) HAT-P-67, based on comparison to the Geneva isochrones. In Section \ref{secMISTtracks} MIST evolutionary tracks showed two equally plausible states: the tail-end of the main sequence or a recently evolved sub-giant.  Figure \ref{fig:OhmicInflate} employs these tracks to assess the re-inflationary impact of \hatpb~could be, assuming that the orbital location has not changed over the system lifetime.  The second panel from the top illustrates a countervailing effect: the decrease in Ohmic dissipation heating efficiency, $\epsilon$, could counteract the increase of stellar flux, assuming the functional form for $\hat{\epsilon}_G$ described by \citet{2018AJ....155..214T}.  Together the effects nearly cancel when computing the inflationary lifetimes, yielding nearly identical curves in the bottom panel.  The third panel recreates the numerically computed curve for $0.5\;M_\mathrm{Jup}$, $T_\mathrm{eq}=$1800 K from \citep{2011ApJ...738....1B}, which should be considered as representative since it was not necessarily tailored to the evolutionary history of \hatpb.  Nevertheless, the similarity of the curve is remarkable since it arrives at approximately the correct planet radius at the right age.  The modeled mass is slightly higher than, but consistent with the RV-measured mass, and would imply a large and increasing mass loss rate over its lifetime.

@@

A doubling of insolation would place the young \hatpb~at a 20\% lower equilibrium temperature of 1600 K.  Perversely, the $\epsilon_G(T_\mathrm{eq})$ may tend to \emph{decrease} over time since the heating efficiency appears to peak at 2.5\% at 1600 K \citep{2018AJ....155..214T}, and reduce to about 1\% at 1900 K.  So inflation timescales may remain relatively unchanged over the planet lifetime, despite a doubling of incident radiation.

@@

>>>>

% This is the version destined for Overleaf on June 7th, 2023.
\title{A large and variable leading tail of Helium in a sub-Saturn undergoing runaway inflation}

@@

A runaway inflation scenario predicts an inevitable and profound mass loss rate for inflated hot Saturns.  As the planet's atmosphere overflows its Roche lobe at an ever-increasing pace, instantaneous mass loss rates may exceed $\dot{M}>10^{13}$ g/s, over 10$\times$ larger than those previously seen in atmospheric escape measurements to date \citep{2022arXiv221116243D}.  Inflated hot Saturns, therefore, make excellent targets for direct measurement of atmospheric escape.

Large mass loss alone does not guarantee detectability.  The ability to detect even immense mass loss hinges on its observability in spectral tracers.  Ly$\alpha$, \ion{He}{1} 10833 \AA, and H$\alpha$ have emerged as the most amenable to detection \citep{2018ApJ...855L..11O,2022arXiv221116243D,2023MNRAS.518.4357O}, but each of these has its own observational limitations.  Ly$\alpha$ can suffer from interstellar medium \ion{H}{1} absorption censoring its low-velocity line core, for example.  Here we focus on \ion{He}{1} 10833 \AA, which offers some advantages.  In particular, metastable Helium's ability to be observed at high spectral resolution from the ground has been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.  \ion{He}{1} has resulted in at least 14 systems with detections, including \object[GJ 3470]{GJ 3470 b} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.  An observational picture of mass loss in hot Saturns appears to be lacking, at least in part due to their intrinsic rarity.

@@

Here we present a multi-year observational campaign searching for atmospheric escape in \object{HAT-P-67 b}, an inflated hot Saturn orbiting an F type sub-giant at an orbital separation of 0.06 AU and a 4.8-day orbital period \citep{2017AJ....153..211Z}.  Its strong insolation ($T_\mathrm{eq}\sim2000\;$K) combined with \object{HAT-P-67 b}'s extremely low surface gravity makes it an exceptional candidate for strong atmospheric mass loss via Roche lobe overflow.  Its $\sim$0.05 g~cm$^{-3}$ density places it below the 0.1 g~cm$^{-3}$ threshold predicted to exhibit runaway inflation by \citet{2023ApJ...945L..36T}.  Figures \ref{fig:instellation} and \ref{fig:ThornFortUpdated} show how much of an outlier \hatpb~ is: large, low mass, and heavily irradiated.

The evolutionary state of HAT-P-67~b offers even more intriguing possibilities.  The host star may be an evolved subgiant, opening the prospect for planetary re-inflation \citep{2022AJ....163...53S, 2022AJ....163..120G, 2023arXiv230306728G} as the planet's insolation increased with the star's rapid luminosity increase in this part of the short-lived HR diagram.  The tidal gravity of the nearby massive host star may amplify mass loss rates even further \citep{2007A&A...472..329E, 2023ApJ...945L..36T}. Its status as a rare inflated Hot Saturn makes HAT-P-67~b a promising laboratory, uniquely suited for testing the runaway inflationary predictions of Ohmic dissipation and XUV photoevaporation.

We assemble both archival, previously published, and new observations of the HAT-P-67 system, chronicled in Section \ref{secObs}.  In Section \ref{secAnalysis}, we refine the stellar and planet properties based on those observations, including distance, radius, and orbit re-analyses.  The Helium excess detection is presented in Section \ref{secHeAnalysis}, with deep dives on the signal's trend with orbital phase.  We use this orbital structure to reconstruct the geometry and mass loss of the escaping Helium exosphere, with both 1D and 3D models (Section \ref{secResults}).  We assess the physical mechanisms (Section \ref{secPhysMech}) giving rise to the escaping atmosphere, weighing the distinctive predictions of XUV irradation and Ohmic dissipation.  Finally in Section \ref{secDiscuss}, we question the assumptions in our approach, discuss the overall congruence of predictions and observations, and highlight some implications for future exosphere studies.

@@

The Habitable Zone Planet Finder Spectrograph \citep[HPF;][]{2012SPIE.8446E..1SM,2014SPIE.9147E..1GM, 2019Optic...6..233M} on the queue-scheduled 10-meter \emph{Hobby-Eberly Telescope} \citep[HET;][]{1998SPIE.3352...34R} operates in the near-IR from $8100-12800~$\AA~ spanning the \textit{z}, \textit{Y}, and \textit{J} bands at spectral resolving power $R=55,000$. The HET fixed-elevation design \citep{2007PASP..119..556S} limits the observability of \object{HAT-P-67} to less than 1 hour ``tracks'' for a fixed range of hour angles before (east track) and after (west track) the star transits the meridian.  Whereas conventional steerable telescopes could conduct continuous point-and-stare observations of \object{HAT-P-67} for hours, HET cannot.  In practice, this limitation means that in-transit and out-of-transit observational phases were rarely possible on the same night.  Instead, we organized the observations into four campaigns to coincide with \object{HAT-P-67 b} transits on the nights beginning 2020 April 27, 2020; May 21, 2020; June 14, 2020; and 2022 April 28.  These campaigns have out-of-transit observations at least one night before and one night after, and often two nights on either side of the transit.  Two more transit snapshots were obtained in 2022 June-July without the accompanying visits immediately before and after.  The in-transit campaigns had up to 14 exposures per HET track, with integration times between 5-8.5 minutes.  We also obtained random-in-phase lower-priority P1-4 reconnaissance observations \citep{2007PASP..119..556S}.  These out-of-transit snapshot observations typically received 4 or fewer individual exposures.  We observed HAT-P-67 with HPF for a total of 41 visits on 39 unique nights, with two of those nights observing both the east and west HET tracks.  The total on-source integration time exceeds 13.8 hours, over $N=152$ individual spectra.  Table \ref{tabHPFLog} summarizes the log of observations.

@@

\subsection{TESS Light Curve}
Previously, \hatpb~transits had only been detected with \emph{HATNet} \citep{2004PASP..116..266B} and followed up with KeplerCam on the FLWO 1.2 m telescope \citep{2017AJ....153..211Z}. These ground-based photometers were not intended to measure weak, long-term stellar variability signals.  We, therefore, examined the \emph{TESS} lightcurves for out-of-transit photometric variability.  The revised precision and continuous coverage of \emph{TESS} can also refine the orbital solution.

@@

We reduced and analyzed the HPF data first for orbit fitting via precision radial velocity (PRV) measurements following the procedures described in \citet{2021AJ....161..173T} and updated for joint lightcurve and RV fits \citep{2022AJ....163..225T}.  Here we included the new TESS lightcurves and the existing Keck HIRES points reported in Table 3 of \citet{2017AJ....153..211Z}.  The HPF points exhibited an RV jitter of $\sim164\;\ms$, a few times larger than the typical Keck HIRES measurements.  So even though the HPF points were more numerous, their marginal value for RV orbit determination was subdued owing to the lower RV information content in the near-IR for this relatively rapidly rotating F star.  Figure \ref{fig:RVfit} shows the fitted RV semiamplitude $K=33_{-15}^{+21}$ \kms---assuming a circular orbit---slightly decreasing the uncertainty compared to the previous $1\sigma$ limit.  Table \ref{tabOrbit} lists the revised properties, with minor differences from the existing values.
\authorcomment1{Quang Tran: please verify and revise.}

@@

\subsection{Velocity substructure}
Figure \ref{fig:centroids} shows the centroid positions of the Helium excess feature.  The bulk velocity drifts of the planet's outflow can be seen with a few conspicuous trends.  The pre-transit phases ($-0.3$ to $+0.03$) show a slight tendency towards blueshift relative to stellar restframe, with an accelerating blueshift from phases $-0.3$ to $-0.15$, albeit with some visit-to-visit scatter.  The post-transit observations ($+0.03$ to $+0.15$) exhibit weak-albeit-significant equivalent widths, as denoted by their smaller marker size.  The centroids dramatically accelerate to larger bulk blueshifts, as low as 10831 \AA.

At the moment of transit midcenter, the centroids exhibit a redshifted absorption relative to the planet restframe. Measurements from April 2022 and May 2020 transits partially overlap in phase coverage while yielding slightly different line centroid locations.  Both time series sequences appear consistent with near-zero bulk restframe velocity, but with significant variability in the line profile substructure.  The May 2020 sequence shows a slight trend towards alignment with the planet restframe.  The April 2020 and June 2020 partial transits sampled phases close to planet egress, with a clear blueshift relative to both the star and planet.

@@

The predicted Helium ionization depends sensitively on the XUV flux \citep{2019ApJ...881..133O}.  Ideally, we would have a measurement of the XUV spectrum of HAT-P-67, but the limited facilities and distance of the source preclude these challenging observations.  Instead, we constructed panchromatic X-ray to visible SEDs by scaling and stitching together synthetic spectra to provide a range of high and low $L_X/L_\mathrm{bol}$.  We chose a 6400 K PHOENIX \citep{husser13} solar metallicity photospheric spectral template with $\log{g}=3.5$, scaled to the solid angle seen by \hatpb.  The F7IV-V \object[tau Boo]{$\tau$ Boo} serves as the closest analog with available synthetic X-ray coronal spectra \citep{2011A&A...532A...6S}.  But \hatp resides closer to the Kraft break than does \object[tau Boo]{$\tau$ Boo}, and the move to hotter F stars may be associated with weakening of the stellar wind and other atmospheric changes leading to the prospect of much lower XUV luminosity for HAT-P-67 \citep{2022ApJ...930....7A}.  So there remains significant uncertainty about the applicability of the \object[tau Boo]{$\tau$ Boo} XUV spectrum to HAT-P-67.  We scaled the synthetic SED of \object[tau Boo]{$\tau$ Boo} retrieved from X-exoplanets \citep{2011A&A...532A...6S} so that the integral of ground-state ionizing photons ($\lambda<504\;$\AA) exhibited $L_X/L_\mathrm{bol} \in [10^{-6}, 10^{-4}]$.  Figure \ref{fig:XUV} shows two conceivable SEDs with these high and low radiation hardnesses.

@@

Equipped with these two SEDs of differing radiation hardness, we explored different mass loss rates and exosphere temperatures.  Figure \ref{fig:pwinds} shows one example model spectrum with $\dot{M} = 2\times 10^{13}$ g/s, and $T_0=14000$~K, with an XUV spectrum possessing $L_x/L_\mathrm{bol}=10^{-5}$.  This mass loss rate would imply a characteristic lifetime of $\sim$2000 Myr.  The model spectrum exhibits approximately the same equivalent width as our HPF observations, making it a hypothetical scenario among a family of partially degnerate solutions.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/pwinds_May2020_14000_2e13Mdot_medXUV_formal_rv copy.png}
    \caption{Simulated 1D Parker wind model of Helium absorption in \hatpb.  The spectrum was generated with a mass loss rate of $2\times10^{13}$ g/s, exosphere gas temperature of 14000 K, and with an XUV luminosity $L_x/L_\mathrm{bol}=10^{-5}$ intermediate between those shown in Figure \ref{fig:XUV}.  The \texttt{p-winds} model yields the \ion{He}{1} 10830 triplet lines (colored dashed lines), with a cumulative feature in black resembling the data, shown as the May 2020 in-transit mean.  Additional 3D velocity dispersion of the gas can explain differences between the line shapes of the model and data.  Overall the 1D models may be too simplistic to represent the inherently 3D structure of the outflow.}
    \label{fig:pwinds}
\end{figure}

@@

Orbtial shear, stellar wind confinement, and preferential mass loss may all conspire to create the conditions of high enough column density to populate the \ion{He}{1} metastable triplet to detectable levels while imbuing the velocity substructure that we see.  Here we interpret the bulk velocity substructure in Figure \ref{fig:centroids} under these gas dynamics mechanisms.  We attribute the pre-transit arc to stellar wind acceleration.  The stellar wind is initially too weak to plow the dense escaping gas until a separation of 60-130 planetary radii ($+0.15$ orbital phase), the inflection point in an enormous bow shock.  Gas at these separations has had enough time to diffuse, both out of the orbital plane and to slightly larger radii shells, in turn lowering the column density along the line-of-sight.  The lower column density provides both fewer Helium atoms to participate in absorption, and less overall NUV shielding, leading to greater fractional ionization and further depopulation of the \ion{He}{1} metastable state.

@@

\begin{equation}
    \dot{M} = \frac{3}{4}\frac{\eta F_\mathrm{XUV}}{G K_t \rho_\mathrm{XUV}}\label{thorn23Mdot}.
\end{equation}
 where $K_t$ is the tidal gravity term, and $\eta \sim 0.4$ comes from \citet{2022A&A...663A.122C}.  Assuming $\log{L_\mathrm{XUV}/L_\mathrm{bol}}=-4.2$, we would expect \hatpb~ to exhibit $\dot{M}\sim10^2$ M$_\oplus$/Gyr. At a current $M\sim95\;$M$_\oplus$ and an accelerating mass loss rate, its lifetime would measure in the $\sim$100 Myr range.  The \citet{2023ApJ...945L..36T} simulations focused on 0.75-1.25 $M_\odot$ host stars; the higher mass $\sim$1.6 $M_\odot$ \hatp would deliver greater tidal gravity for a given semi-major axis, and possibly lower $\log{L_\mathrm{XUV}/L_\mathrm{bol}}$ compared to more active G stars.  Nevertheless, we can project the time series trends in their Figures 3 and 4 to recreate a qualitative history and fate of \hatpb~ under the assumptions of XUV photoevaporation.  Broadly, the planet would have started with an initial mass up to 50$\%$ greater than at present, with $R\sim\;$1.4 $R_\mathrm{Jup}$, for an initial density of $\sim$0.2 g/cm$^-3$.  It would lose mass at a rate of tens of Earth masses per Gyr for its 1 Gyr lifetime, expanding modestly until it reaches the critical $\sim$0.1 g/cm$^{-3}$ threshold, at which point the mass loss rate increases to its current value.  It will last only tens of Myr in its current state before losing almost all of its envelope and settling as a $5-15$ M$_\oplus$ core with a final radius of 0.2$-$0.3 R$_\mathrm{Jup}$.

\subsection{Ohmic Dissipation-driven mass loss}
The Ohmic dissipation mechanism made two key observable predictions.  Anomalous heating efficiency, $\epsilon$, should initially increase as a function of equilibrium temperature, then degrade as irradiation exceeds equilibrium temperatures of about 2000 K.  Second, hot Saturns ($\lesssim0.5 M_\mathrm{Jup}$) should undergo runaway evaporation, whereas hot Jupiters ($\gtrsim 1 M_\mathrm{Jup}$) should reach stable equilibrium---albeit inflated---radii after Gyr timescales.  Both of these outcomes predated data that could validate them, and these outcomes differ from the behavior of other heating mechanisms, such as photoevaporation or tides.

@@

A few caveats complicate the unambiguous causal interpretation of Ohmic dissipation.  First, the scaling law arguments that produced Equation \ref{eqInflate} were only proposed as coarse estimates, with numerical simulations needed to quantify inflation timescales for individual systems.  Accordingly, a $10\times$ higher inflation timescale of 50 Myr---allowed by the coarse scaling law---would yield $\dot{M}_\mathrm{infl}\sim10^{14}$ g/s, still a very large mass loss rate, and only a factor of 5 away from the baseline 1D model.  Factors of a few uncertainties in the Ohmic dissipation efficiency $\epsilon$ may also contribute.

@@

\citet{2017AJ....153..211Z} estimated that \hatpb~received about twice the incident flux as a Zero Age Main Sequence (ZAMS) HAT-P-67, based on comparison to the Geneva isochrones. In Section \ref{secMISTtracks} MIST evolutionary tracks showed two equally plausible states: the tail-end of the main sequence or a recently evolved sub-giant.  Figure \ref{fig:OhmicInflate} employs these tracks to quantify the prospects for re-inflation  of \hatpb.

We further assume that the orbital location has not changed over the system lifetime, that the planet's response to anomalous heating stimulus is instantaneous, and that the anomalous heating efficiency $\hat{\epsilon}_G$ peaks at $T_\mathrm{eq}=1750$ K as derived by \citet{2018AJ....155..214T}.

We conclude that reinflation likely does not have a significant effect on the evolution of HAT-P-67~b: inflation timescales remain relatively unchanged over the planet lifetime, despite a 50\% spike in incident radiation in the sub-giant scenario.  The reason is subtle, as we describe next.

The second panel from the top illustrates a countervailing effect: the increase in flux actually triggers a decrease in anomalous heating efficiency, $\epsilon$. The greater stellar energy couples into the planet less effectively than the weaker radiation did, roughly balancing out. Together the effects nearly cancel when computing the inflationary lifetimes, yielding nearly identical curves in the bottom panel: a secularly evolving main sequence history gives almost the same inflation timescale as a rapidly increasing subgiant.  It is important to emphasize that the anomalous heating efficiency found by \citet{2018AJ....155..214T} is agnostic to what the actual heating mechanism is: either XUV Irradiation or Ohmic Dissipation both have to obey the trend of $\hat{\epsilon}_G(T_\mathrm{eq})$.

The third panel recreates the numerically computed curve for $0.5\;M_\mathrm{Jup}$, $T_\mathrm{eq}=$1800 K from \citep{2011ApJ...738....1B}, which should be considered as representative since it was not necessarily tailored to the evolutionary history of \hatpb.  Nevertheless, the similarity of the curve is remarkable since it arrives at approximately the correct planet radius at the right age for about the right initial mass.

@@

@@

    GS acknowledges support provided by NASA through the NASA Hubble Fellowship grant HST-HF2-51519.001-A awarded by the Space Telescope Science Institute, which is operated by the Association of Universities for Research in Astronomy, Inc., for NASA, under contract NAS5-26555.
